---
title: "Case Increase Trends by County"
author: "Kara Haberstock Tanoue"
date: "7/05/2020"
output: 
  html_document:
    theme: paper
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(plotly)
library(gapminder)
library(hrbrthemes)
library(RCurl)
library(tidyverse)
```

## 7-Day Average Daily Case Increases in Arizona Counties

Average daily case increases are on an upward trend in most counties in Arizona.
  
Note that typically it takes 10-14 days to see the effects of major policy changes or events. This is due to both the incubation period of COVID-19 (median of 4-5 days between exposure and symptom onset according to the CDC) and the reporting lag for new case data (typically 4-7 days according to ADHS).  
  
#### Key Dates in Arizona  

**03/15**: Schools Close  
**03/31**: Stay at Home Order goes into effect  
**04/23**: Testing Expands  
**05/15**: Stay at Home Order expires  
**05/25**: Memorial Day  


```{r echo=FALSE}
NYTcsvpull<-getURL("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
CountyCOVIDCases<-data.frame(read.csv(text = NYTcsvpull))
AZCountyCOVID<-subset(CountyCOVIDCases, state == "Arizona")
AZCountyCOVID<-subset(AZCountyCOVID, county != "Unknown")
AZCountyCOVID$date<-as.Date(AZCountyCOVID$date)

#calculate avg daily case increase
AZCountyCOVID<-AZCountyCOVID  %>% 
  group_by(county) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(D7AVGINC=(cases-lag(cases,7))/7)

AZCountyCOVID<-AZCountyCOVID %>%
  ungroup(county)

#subset to no NA
AZCountyCOVID<-subset(AZCountyCOVID, AZCountyCOVID$date >= as.Date('2020-3-9'))
AZCountyCOVIDinc<-AZCountyCOVID[!is.na(AZCountyCOVID$D7AVGINC),]

countyincfig<-ggplot(AZCountyCOVIDinc, aes(x=date, y=D7AVGINC, group=county, color=county, text = paste0(format.Date(date,"%m/%d"), ": ",format(D7AVGINC, digits = 2))), size=2, alpha=0.7) +
  geom_point()+
  facet_wrap(county ~ ., ncol = 1, scales = "free_y")+
  theme_ipsum(base_family = "Arial Narrow")+
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), legend.position = 'none')+
  scale_x_date(name="Date", date_breaks = "2 weeks", date_minor_breaks = "1 day", date_labels = "%m/%d")+
  scale_y_continuous(name = "Case Increase\n")+
  guides(color= FALSE, size= FALSE, alpha= FALSE)

ggplotly(countyincfig, width= 800, height= 2000, tooltip = "text") %>%
  layout(hovermode = "x unified") %>%
  style(countyincfig, mode="markers+lines")

```

### Data & Methods:

County data on COVID-19 case counts are drawn from the New York Times COVID-19 Data Repsitory, accessed at [https://github.com/nytimes/covid-19-data](https://github.com/nytimes/covid-19-data). The 7-Day Average Daily Case Increase variable is a derived variable obtained by averaging the new cases added each day over a rolling 7-day window. A 7-day window was used as there are strong day of the week effects in data reporting. These data were last updated on 07.04.2020 
  
**Data analyzed and visualized by [Kara Haberstock Tanoue](https://www.karatanoue.com).**

### Other Dashboards Worth Following:

**[COVID Risk Levels](https://globalepidemics.org/key-metrics-for-covid-suppression/)**: Provides county-level risk levels based on population-normalized case growth  
**[COVID Act Now](https://covidactnow.org)**: Provides a model of infection rates and key metrics for states and counties across the U.S.  
**[COVID Racial Data Tracker](https://covidtracking.com/race)**: Provides data on race & ethnicity for COVID-19 cases across the U.S. and tracks racial and ethnic disparities in cases and deaths. 
