---
title: "Trends in COVID 19 Case Increases & Key Hospital Metrics"
author: "Created by Kara Haberstock Tanoue"
date: "7/06/2020"
output: 
  html_document:
    theme: paper
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(plotly)
library(gapminder)
library(hrbrthemes)
```

## 7-Day Average Daily Case Increases in Arizona & Pima County

Average daily case increases in Arizona and Pima County have rapidly increased since Memorial Day Weekend. Tuesday, June 23, had a record single-day case increase in Pima County (502). Wednesday, July 1, saw a record single-day case increase in Arizona (4,828).
  
Note that typically it takes 10-14 days to see the effects of major policy changes or events. This is due to both the incubation period of COVID-19 (median of 4-5 days between exposure and symptom onset according to the CDC) and the reporting lag for new case data (typically 4-7 days according to ADHS).  
  
Average daily cases increases began to reliably increase beginning on May 29, which was 14 days following the expiration of stay at home orders in Arizona.   


#### Key Dates in Arizona  

**03/15**: Schools Close  
**03/31**: Stay at Home Order goes into effect  
**04/23**: Testing Expands  
**05/15**: Stay at Home Order expires  
**05/25**: Memorial Day  

```{r echo=FALSE}
RAWAZPIMACOVID<-read.csv(file='COVID AZ PIMA Cases Simple.csv')
AZPIMACOVID<- data.frame(RAWAZPIMACOVID)
#convert excel formatting to date
AZPIMACOVID$CaseDate<-as.Date(AZPIMACOVID$Date,origin = "1899-12-30")
attach(AZPIMACOVID)
#summary(AZPIMACOVID)

update_geom_defaults("text", list(size=10))
casefig<-ggplot(data=AZPIMACOVID, aes(x=CaseDate, y=AvgIncrease), size=2, alpha=0.7)+
  geom_point(aes(y = AZ7DayAvgInc, color= "AZ 7-Day Avg Increase", text = paste0(format.Date(CaseDate,"%m/%d"),": ",format(AZ7DayAvgInc, digits = 2))))+
  geom_point(aes(y = Pima7DayAvgCaseInc, color= 'Pima County 7-Day Avg Increase', text = paste0(format.Date(CaseDate,"%m/%d"),": ", format (Pima7DayAvgCaseInc, digits = 2))))+
  geom_vline(xintercept = as.numeric(as.Date('2020-3-15')), linetype='dashed', color='#FD6035', size=1)+
  geom_text(data = AZPIMACOVID, aes(x=as.Date('2020-3-15'), y= AZ7DayAvgInc[length(AZ7DayAvgInc)]), label= 'Schools Close', color= '#FFFFFF', size=1, family= 'Arial Narrow', fontface='plain')+
  geom_vline(xintercept = as.numeric(as.Date('2020-3-31')), color='#DE3202', size=1)+
  geom_text(data = AZPIMACOVID, aes(x=as.Date('2020-3-31'), y= AZ7DayAvgInc[length(AZ7DayAvgInc)]), label= 'Stay at Home Order', color= '#FFFFFF', size=3, family= 'Arial Narrow', fontface='plain')+
  geom_vline(xintercept = as.numeric(as.Date('2020-4-23')), linetype= 'dashed', color= '#0E7C7B', size = 1)+
  geom_text(data = AZPIMACOVID, aes(x=as.Date('2020-4-23'), y= AZ7DayAvgInc[length(AZ7DayAvgInc)]), label= 'Expanded Testing', color= '#FFFFFF', size=1, family= 'Arial Narrow', fontface='plain')+
  geom_vline(xintercept = as.numeric(as.Date('2020-5-15')), color='#DE3202', size= 1)+
  geom_text(data = AZPIMACOVID, aes(x=as.Date('2020-5-15'), y= AZ7DayAvgInc[length(AZ7DayAvgInc)]), label= 'Order End', color= '#FFFFFF', size=1, family= 'Arial Narrow', fontface='plain')+
  geom_vline(xintercept = as.numeric(as.Date('2020-5-25')), linetype = 'dashed', color= '#cc8f00', size= 1)+
  geom_text(data = AZPIMACOVID, aes(x=as.Date('2020-5-25'), y= AZ7DayAvgInc[length(AZ7DayAvgInc)]), label= 'Memorial Day', color= '#FFFFFF', size=1, family= 'Arial Narrow', fontface='plain')+
  scale_color_manual(values=c('#1FFFFF','#60079C'))+
  theme_ipsum(base_family = 'Arial Narrow')+
  theme(legend.position = 'bottom')+
  scale_x_date(name="Major Dates", breaks = c(as.Date('2020-3-15'), as.Date('2020-3-31'), as.Date('2020-4-23'), as.Date('2020-5-15'), as.Date('2020-5-25')), date_minor_breaks = "1 day", date_labels = "%m/%d")+
  scale_y_continuous(name = "Case Increase", breaks = seq(0,3500,500))+
  scale_alpha(guide="none")+
  scale_size(guide="none")+
  guides(color=guide_legend(title=NULL))


ggplotly(casefig, width= 800, height= 600, tooltip = "text") %>%
  layout(hovermode = "x unified", legend= list(orientation= "h", x=0.2, y=-0.1)) %>%
  style(casefig, mode="markers+lines", opacity=0.7)
```

The 7-Day Average Daily Case Increase variable is a derived variable obtained by averaging the new cases added each day over a rolling 7-day window. A 7-day window was used as there are strong day of the week effects in data reporting.

## Tests per Positive 

This is a re-calculation of the case-to-test ratio based on guidance from the [COVID Tracking Project](https://covidtracking.com/blog/the-other-covid-19-metric). The tests per positive metric tells us how many tests we need to perform in order to find a positive case. Higher is better and indicates more testing. Currently, 1 in 4 tests in Arizona are returning a positive case, and 1 in 6 tests in Pima County are returning a positive case. In countries that have controlled the COVID-19 outbreak well, such as Thailand, 1 in 1,000 tests are rturning a positive case. We have a long ways to go.

```{r echo=FALSE}
CTRfig<-ggplot(data=AZPIMACOVID, aes(x=CaseDate, y=CaseToTest), size=2, alpha=0.7)+
  geom_point(aes(y = AZCTR, color= "AZ 7-Day Avg Case-Test Ratio", text = paste0(format.Date(CaseDate,"%m/%d"),": ",format(AZCTR, digits = 2))))+
  geom_point(aes(y = PIMACTR, color= 'Pima County 7-Day Avg Case-Test Ratio', text = paste0(format.Date(CaseDate,"%m/%d"),": ", format (PIMACTR, digits = 2))))+
  scale_color_manual(values=c('#1FFFFF','#60079C'))+
  theme_ipsum(base_family = 'Arial Narrow', base_size= 11.5)+
  theme(legend.position = 'bottom')+
  scale_x_date(name="Date", date_breaks = "2 weeks", date_minor_breaks = "1 day", date_labels = "%m/%d")+
  scale_y_continuous(name = "Tests Per Positive | More Testing >>>", breaks = seq(0,40,5))+
  scale_alpha(guide="none")+
  scale_size(guide="none")+
  guides(color=guide_legend(title=NULL))

ggplotly(CTRfig, width= 800, height= 600, tooltip = "text") %>%
  layout(hovermode = "x unified", legend= list(orientation= "h", x=0.2, y=-0.2)) %>%
  style(CTRfig, mode="markers+lines", opacity=0.7)

```

If the tests per positive is decreasing, this indicates **insufficient testing to keep up with the outbreak**.  
  
## Key Hospital Metrics

Hospital COVID-19 specific metrics have been rapidly trending up in the last 3 weeks across Arizona. Hospital in-patient bed usage is currently at 84% and ICU bed usage is at 89% statewide (note that beds in use include both COVID and non-COVID patients).

```{r echo=FALSE}
detach(AZPIMACOVID)


#split out hospital data to nix NAs
HOSPAZPIMA<-AZPIMACOVID[!is.na(AZPIMACOVID$AZinHosp),]
HOSPAZPIMA<-data.frame(HOSPAZPIMA)
attach(HOSPAZPIMA)
#summary(HOSPAZPIMA)



#Plotting Hospital Data-- import necessary libraries


hospfig<-ggplot(data = HOSPAZPIMA, aes(x = CaseDate, y = Patients/Beds), size=2, alpha=0.7) + 
  geom_point(aes(y = AZinHosp, color= "AZ Inpatient", text = paste0(format.Date(CaseDate,"%m/%d"),": ",AZinHosp, " Inpatient")))+
  geom_point(aes(y = AZinED, color= 'AZ Emergency Dept', text = paste0(format.Date(CaseDate,"%m/%d"),": ",AZinED, " ED")))+
  geom_point(aes(y = AZinICU, color= 'AZ ICU Beds', text = paste0(format.Date(CaseDate,"%m/%d"),": ",AZinICU, " ICU")))+
  geom_point(aes(y = AZwithVent, color= 'AZ on Ventilator', text = paste0(format.Date(CaseDate,"%m/%d"),": ",AZwithVent, " Vents")))+
  scale_color_manual(values=c('#1FFFFF','#0E7C7B','#60079C','#DE3202'))+
  theme_ipsum(base_family = 'Arial Narrow', base_size= 11.5)+
  theme(legend.position = 'bottom')+
  scale_x_date(name="Date", date_breaks = "1 week", date_minor_breaks = "1 day", date_labels = "%m/%d")+
  scale_y_continuous(name = "Patients or Beds", breaks = seq(0,3200,200))+
  scale_alpha(guide="none")+
  scale_size(guide="none")+
  guides(color=guide_legend(title=NULL))


ggplotly(hospfig, width= 800, height= 600, tooltip = "text") %>%
  layout(hovermode = "x unified", legend= list(orientation= "h", x=0.1, y=-0.2)) %>%
  style(hospfig, mode="markers+lines", opacity=0.7)
```


All hospital metrics are collected by ADHS through the EMResource platform. The data reflect hospital reported bed usage and resource availability at 5:00 PM the preceding day.

### Data & Methods:

All data on COVID-19 case counts and hospital metrics are drawn from the Arizona Department of Health Services (ADHS) COVID-19 Data Dashboard, accessed at [https://www.azdhs.gov/preparedness/epidemiology-disease-control/infectious-disease-epidemiology/covid-19/dashboards/index.php](https://www.azdhs.gov/preparedness/epidemiology-disease-control/infectious-disease-epidemiology/covid-19/dashboards/index.php). These data were last updated at 9:00 AM on 07.06.2020 
  
**Data analyzed and visualized by [Kara Haberstock Tanoue](https://www.karatanoue.com).**

### Other Dashboards Worth Following:

**[COVID Risk Levels](https://globalepidemics.org/key-metrics-for-covid-suppression/)**: Provides county-level risk levels based on population-normalized case growth  
**[COVID Act Now](https://covidactnow.org)**: Provides a model of infection rates and key metrics for states and counties across the U.S.  
**[COVID Racial Data Tracker](https://covidtracking.com/race)**: Provides data on race & ethnicity for COVID-19 cases across the U.S. and tracks racial and ethnic disparities in cases and deaths. 